Towards a framework for reasoning about stateful compiler IRs using coinductive reasoning.

### 2024-06-25 paper.tex
fixes

Lopes et al.~propose \emph{translation validation} as a way to scale
formal verification~\cite{lopesAlive2BoundedTranslation2021}.
Their implementation, called Alive, has found wide adoption in the
development of LLVM. Given a declarative description of an optimization
as a source and target pattern in LLVM IR, the tool uses an SMT solver
to automatically determine whether the optimized pattern exhibits only
behaviours of the original program.

@@

For example, Xia et al.~used interaction trees, a coinductive type, in
their Coq formalization of recursive and impure programs~\cite{xiaInteractionTreesRepresenting2020}.
Similarly, MLIR has an
``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
centred on finite state machines. Such automata are understood to be
coinductive in nature, so mechanizing the semanantics of this dialect in
Lean will likely require coinduction.

@@

Here, \texttt{y} is obtained as the bitwise and of the argument
\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
sense, but this style of definition is common in hardware
representations, where this is interpreted as: ``the value of \texttt{y}
in the \emph{current} clock cycle is computed by taking the bitwise and
of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''.~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles. See https://circt.llvm.org/ for more details} To formally model this, we could denote variables as infinite streams of bitvectors,
which gives us the value at every possible future clock cycle. Another
example of a dialect with graph regions is the FSM, or Finite State
Machine, dialect.

@@

However, coinduction is actually a very good fit for this problem as well.
Therefore, I'd like to use my library for coinductive types to define a
coinductive monad which supports computable infinite behaviours, drawing
inspiration from the interaction tree library for coq.~\cite{xiaInteractionTreesRepresenting2020}
This addresses a
weakness in LeanMLIR while simultaneously giving a nice case study for
the usability of the coinduction library.

>>>>

Lopes\etal{} propose \emph{translation validation} as a way to scale
formal verification~\cite{lopesAlive2BoundedTranslation2021}.
Their implementation, called Alive, has found wide adoption in the
development of LLVM. Given a declarative description of an optimization
as a source and target pattern in LLVM IR, the tool uses an SMT solver
to automatically determine whether the optimized pattern exhibits only
behaviours of the original program.

@@

For example, Xia\etal{} used interaction trees, a coinductive type, in
their Coq formalization of recursive and impure programs~\cite{xiaInteractionTreesRepresenting2020}.
Similarly, MLIR has an
``fsm'' dialect~\footnote{https://circt.llvm.org/docs/Dialects/FSM/}
centred on finite state machines. Such automata are understood to be
coinductive in nature, so mechanizing the semanantics of this dialect in
Lean will likely require coinduction.

@@

Here, \texttt{y} is obtained as the bitwise and of the argument
\texttt{x} and \texttt{y} itself. As a program, this doesn't make much
sense, but this style of definition is common in hardware
representations, where this is interpreted as: ``the value of \texttt{y}
in the \emph{current} clock cycle is computed by taking the bitwise and
of the argument \texttt{x} and the \emph{previous} value of \texttt{y}''~\footnote{This example is simplified, in reality you'd have to specify a register instruction to control when a value is stored across clock cycles. See https://circt.llvm.org/ for more details}.
To formally model this, we could denote variables as infinite streams of bitvectors,
which gives us the value at every possible future clock cycle. Another
example of a dialect with graph regions is the FSM, or Finite State
Machine, dialect.

@@

However, coinduction is actually a very good fit for this problem as well.
Therefore, I'd like to use my library for coinductive types to define a
coinductive monad which supports computable infinite behaviours, drawing
inspiration from the interaction tree library for Coq~\cite{xiaInteractionTreesRepresenting2020}.
This addresses a
weakness in LeanMLIR while simultaneously giving a nice case study for
the usability of the coinduction library.
